---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)



if (FALSE) {
  covr::report(covr::package_coverage(
    line_exclusions = list()
  ))
}

library(rmonocypher)
set.seed(1)
```

# `{rmonocypher}`: simple encryption tools for R

<!-- badges: start -->
[![R-CMD-check](https://github.com/coolbutuseless/rmonocypher/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/rmonocypher/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{rmonocypher}` provides easy-to-use tools for encrypting data in R, 
based on the [`monocypher`](https://monocypher.org/) library.

The key features are:

* Seamless encryption with many R functions using a `connection`
* Easy encryption of data and strings 
    * Using ['Authenticated Encryption with Additional Data'](https://en.wikipedia.org/wiki/Authenticated_encryption#Authenticated_encryption_with_associated_data_(AEAD)) (AEAD)
* Generate encryption keys from easier-to-remember pass-phrases
    * Using ['Argon2'](https://en.wikipedia.org/wiki/Argon2) for password-based key derivation


## What's in the box

* `cryptfile()` is a connection for reading/writing encrypted data.  
    * This connection automatically encrypts/decrypts data to file and can be used in any 
      function which supports connections 
      e.g. `saveRDS()`, `write.csv()`, `png::writePNG()` etc
* `mc_encrypt()` and `mc_decrypt()` are for encrypting and decrypting raw vectors
  and strings
* `argon2()` derives random bytes (e.g. a `key`) from a pass-phrase
* `isaac()` is a cryptographic RNG for generating random bytes


## Installation

You can install the development version of rmonocypher from [GitHub](https://github.com/coolbutuseless/rmonocypher) with:

``` r
# install.packages("devtools")
devtools::install_github("coolbutuseless/rmonocypher")
```

# Using the `cryptfile()` connection

Any function which supports reading/writing with a connection, now supports 
automatic encryption/decrytion using the `cryptfile()` connection in this package.

Many R functions support connections e.g. `read.csv()`, `saveRDS()`, 
`serialize()`, `png::writePNG()`, and these can all seamlessly use the `cryptfile()` connection.

The output can only be recovered using the same `key` which was 
used for encryption.

```{r}
robj <- head(mtcars)
path <- tempfile()
key  <- argon2("my secret")
saveRDS(robj, cryptfile(path, key))

readRDS(cryptfile(path, key))
```

Attempting to use an incorrect key will not work:

```{r, error=TRUE}
readRDS(cryptfile(path, "wrong key"))
```

Attempting to read the file without decrypting it will not work:

```{r, error = TRUE}
readRDS(path)
```

### More examples using `cryptfile()` connection

```{r eval=FALSE}
# Saving R objects
saveRDS(robj, cryptfile(path, key))
readRDS(cryptfile(path, key))

# CSV files
write.csv(iris, cryptfile(path, key))
read.csv(cryptfile(path, key))

# PNG files
png::writePNG(image, cryptfile(path, key))
png::readPNG(cryptfile(path, key))

# sink()
sink(cryptfile(path, key))
...
sink()

# Text
writeLines(my_diary, cryptfile(path, key))
readLines(cryptfile(path, key))

# cat()
cat(data, cryptfile(path, key))
```



# Simple encrypt/decrypt of raw data and strings

`mc_encrypt()` and `mc_decrypt()` are functions for directly encrypting 
strings and raw vectors

```{r}
library(rmonocypher)

# Plain text. Can be string or raw bytes
dat <- "Hello #RStats"

# Encrypt 
enc <- mc_encrypt(dat, key = "my secret")

# the encrypted data
enc

# Decrypt using the same key
mc_decrypt(enc, key = "my secret", type = 'string')
```

# Encryption keys

The `key` for encryption may be one of:

* A 32-byte raw vector
* A 64-character hexadecimal string
* Any other length string is also acceptable.  In this case, `argon2()` will be used 
  to derive a 32-byte key from this text.  Note: this method will use a 
  default, fixed salt which will reduce the strength of this encryption when 
  facing a skilled adversary.

The key may be set explicitly when the function is called, but can also be set
globally for the current session using `options(MONOCYPHER_KEY = "...")`

Some valid keys for encryption are:

```{r}
# Random raw bytes
key <- as.raw(sample(0:255, 32, TRUE))

# Random raw bytes from isaac()
key <- isaac(32)

# 64-character hexadecimal string
key <- "82febb63ac2ab2a10193ee40ac711250965ed35dc1ce6a7e213145a6fa753230"

# Output from argon2() using an explicit salt
key <- argon2("my secret", salt = "cefca6aafae5bdbc15977fd56ea7f1eb")

# Output from argon2() using an random bytes
key <- argon2("my secret", salt = isaac(16))
```


# Argon2: Password-based Key Derivation

Argon2 is a resource intensive password-based key derivation scheme.

Use `argon2()` to generate random bytes for keys from a pass-phrase.  Defend 
against attackers using rainbow tables by providing extra bytes of `salt`.

```{r echo = FALSE}
set.seed(2024)
```

```{r}
# When no salt is provided, a salt will be 
# derived internally from the pass-phrase.  This is convenient, but 
# not a great security practice (depending on your expected attacker)
argon2("my secret")

# Use text as the salt
argon2("my secret", salt = "salt and vinegar")

# Use a 32-character hexadecimal string as the salt
argon2("my secret", salt = "cefca6aafae5bdbc15977fd56ea7f1eb")

# Use random bytes for the salt
argon2("my secret", salt = as.raw(sample(0:255, 16, TRUE)))

# Use 'isaac()' to source random bytes for the salt
argon2("my secret", salt = isaac(16))
```



# Additional data

**Note:** This is an advanced topic, and not essential for regular use of the encryption 
tools in this package.

This package uses ['Authenticated Encryption with Additional Data (AEAD)'](https://en.wikipedia.org/wiki/Authenticated_encryption#Authenticated_encryption_with_associated_data_(AEAD)) 
to encrypt messages.

Additional data (or 'associated data') may be included in the encryption
process. This additional data is part of the message authentication, but 
not part of the encrypted message.

A common way additional data is used is for encrypted data which has an 
unencrypted header containing meta-information.  This header must be readable
before the message is decrypted, but also must not allow tampering.

An example of the use of *additional data* is addressing an encrypted message to a particular
recipient.

The address on the envelope must be readable to allow delivery, but the message
inside needs to remain confidential.  

In this case, the *address* is the *additional data* -
it is sent with the data, but not encrypted.  Because the address forms
part of the message authentication, any modification of the address will 
prevent the authentication of the encrypted payload.

**In the following example**, a message for Judy is encrypted, and the address
on the envelope is used as the *additional data*.

```{r}
# Using additional data to encrypt a message
key      <- argon2("my secret key2")
message  <- 'Meet me in St Louis'
address  <- 'To: Judy'
enc      <- mc_encrypt(message, key, additional_data = address)

# Package the additional data and deliver to recipient
letter <- list(address = address, message = enc)
letter

# Recipient decodes message. 
# If envelope or contents are tampered with, message decryption will fail.
mc_decrypt(letter$message, key = key, type = 'string', additional_data = letter$address)
```

If a malicious third-party tampers with the address, then the message cannot
be authenticated.  E.g. if the letter is altered as if it were being sent
to "Sandra", then decryption will fail:

```{r error = TRUE}
letter$address <- "To: Sandra"
mc_decrypt(letter$message, key = key, type = 'string', additional_data = letter$address)
```




# Technical Notes

* The file structure is:
    * `[nonce] [frame] [frame] ... [frame]`
        * `[nonce]` = 24 bytes
        * A `[frame]` = `[payload size] [mac] [payload]`
            * `[payload size]` = 8 bytes
            * `[mac]` = 16 bytes
            * `[payload]` is `payload size` bytes
* The nonce is created internally using random bytes from the ISAAC random number
  generator
* There are no magic bytes pre-pended to the data to identify it as encrypted.
* In general when encrypting data using Authenticated Encryption:
    * these should be kept secret:
        * the original data (obviously!)
        * the encryption key.
    * these elements do not need to be kept secret:
        * MAC - message authentication code
        * Number of bytes
        * Nonce
        * Salt (if `argon2()` is being used to derive `key` from a pass-phrase)



#### Built-in libraries

This package uses the [monocypher](https://monocypher.org) encryption library
v4.0.2 to provide 'authenticated encryption with additional data' (AEAD) 
and Argon2 password-based key derivation.

The cryptographic RNG is ISAAC.  Code is in the public domain, and 
available from the [Bob Jenkins' (author) homepage](https://burtleburtle.net/bob/rand/isaacafa.html).

