---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)



if (FALSE) {
  covr::report(covr::package_coverage(
    line_exclusions = list()
  ))
}

library(rmonocypher)
set.seed(1)

if (FALSE) {
  # Check for functions which do not have an @examples block for roxygen
  system("grep -c examples man/*Rd", intern = TRUE) |> 
    grep(":0$", x = _, value = TRUE)
}


if (FALSE) {
  covr::report()
}



if (FALSE) {
  pkgdown::build_site(override = list(destination = "../coolbutuseless.github.io/package/rmonocypher"))
}
```

# Encryption for R Data

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
![](https://img.shields.io/badge/dependencies-zero-blue.svg)
[![CRAN](https://www.r-pkg.org/badges/version/rmonocypher)](https://CRAN.R-project.org/package=rmonocypher)
[![R-CMD-check](https://github.com/coolbutuseless/rmonocypher/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/rmonocypher/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{rmonocypher}` provides a simple, high-level interface for easily encrypting R objects using
a strong, modern cryptographic technique.

The key use-case this package addresses:

    I want to easily encrypt and save data to a public location 
    (e.g. shared drive, cloud drive, etc) which only I can decrypt.


## What's in the box

* `decrypt()`/`encrypt()` read/write encrypted R objects to file 
* `argon2()` derives encryption keys from passwords
* `rbyte()` generates secure random bytes using your operating system's [CSPRNG](https://en.wikipedia.org/wiki/Cryptographically_secure_pseudorandom_number_generator).
  
#### Technical *bona fides* 

* Cryptographic primitives are provided by the included [`monocypher`](https://monocypher.org/) library (v4.0.2)
* Encryption method is [XChaCha20-Poly1305](https://en.wikipedia.org/wiki/ChaCha20-Poly1305) 
  which combines ChaCha20 stream cipher (extended nonce variant) with Poly1305 message authentication.
* Encryption process follows RFC 8439 ['Authenticated Encryption with Additional Data (AEAD)'](https://en.wikipedia.org/wiki/Authenticated_encryption#Authenticated_encryption_with_associated_data_(AEAD)) 
* Key derivation uses [Argon2 password-based key derviation](https://en.wikipedia.org/wiki/Argon2).
* All random bytes are sourced from [Cryptographically secure pseudo-random number generators (CSPRNG)](https://en.wikipedia.org/wiki/Cryptographically_secure_pseudorandom_number_generator).  

## Installation

This package can be installed from CRAN

``` r
install.packages('rmonocypher')
```

You can install the latest development version from
[GitHub](https://github.com/coolbutuseless/rmonocypher) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/rmonocypher')
```

Pre-built source/binary versions can also be installed from
[R-universe](https://r-universe.dev)

``` r
install.packages('rmonocypher', repos = c('https://coolbutuseless.r-universe.dev', 'https://cloud.r-project.org'))
```

## Read/write data to an encrypted file

Encrypt any R object and save to file.

```{r eval = FALSE}
encrypt(mtcars, dst = "SharedDrive/mydata.dat", key = "mykey")
```

Then decrypt the object using the same key.

```{r eval = FALSE}
decrypt(src = "SharedDrive/mydata.dat", key = "mykey")
```


For more details on how passwords are used to derive encryption keys, 
and for other ways of supplying and generating keys see the Vignette: [Encryption Keys](https://coolbutuseless.github.io/package/rmonocypher/articles/encryption-keys.html).



## Vignettes

* [Encryption Keys](https://coolbutuseless.github.io/package/rmonocypher/articles/encryption-keys.html)
    * Generating encryption keys from passwords with `argon2()`
    * Using random bytes as the encryption key
    * Using hexadecimal string as the encryption key
* [Technical Notes](https://coolbutuseless.github.io/package/rmonocypher/articles/technical-notes.html)
    * Background on the encryptiong techniques used
* [Using Additional Data](https://coolbutuseless.github.io/package/rmonocypher/articles/additional-data.html)
    * Advanced technique which is not needed for regular use of this package.
    * Details on using *additional data* 
    * Worked example 








