---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)



if (FALSE) {
  covr::report(covr::package_coverage(
    line_exclusions = list()
  ))
}

library(rmonocypher)
set.seed(1)
```

# `{rmonocypher}`: simple encryption tools for R

<!-- badges: start -->
[![R-CMD-check](https://github.com/coolbutuseless/rmonocypher/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/rmonocypher/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{rmonocypher}` provides easy-to-use tools for encrypting data in R, 
based on the [`monocypher`](https://monocypher.org/) library.

#### Features

* Seamless encryption with many R functions using a `connection`
* Easy encryption of data and strings 
    * Using ['Authenticated Encryption with Additional Data'](https://en.wikipedia.org/wiki/Authenticated_encryption#Authenticated_encryption_with_associated_data_(AEAD)) (AEAD)
* Generate encryption keys from easier-to-remember pass-phrases
    * Using ['Argon2'](https://en.wikipedia.org/wiki/Argon2) for password-based key derivation
* A secure source of random bytes generated from the operating system's entropy sources
* Key sharing using *Shamir's Secret Sharing* 
* General-purpose cryptographic hashing using ['blake2b'](https://en.wikipedia.org/wiki/BLAKE_(hash_function)#BLAKE2) hash function
* Read/write encrypted objects to file

#### Motivating example

Use `{rmonocypher}` to write R data to an encrypted file on a shared drive, 
dropbox or cloud.  

The data is not readable by others, but easily recoverable if the `key` is known.

```{r eval=FALSE}
saveRDS(results, cryptfile("ShareDrive/results.rds", key = "#RsTaTs123!"))

readRDS(cryptfile("ShareDrive/results.rds", key = "#RsTaTs123!"))
```


## What's in the box

* `cryptfile()` is a connection for reading/writing encrypted data.  
    * This connection automatically encrypts/decrypts data to file and can be used in any 
      function which supports connections 
      e.g. `saveRDS()`, `write.csv()`, `png::writePNG()` etc
* `encrypt()` and `decrypt()` are for encrypting and decrypting raw vectors
  and strings
* `argon2()` derives encryption keys from pass-phrases
* `create_public_key()` and `create_shared_key()` can be used to perform key 
  exchange over an insecure channel (i.e. Public Key Cryptography)
* `rcrypto()` is a cryptographic RNG for generating random bytes using the 
  operating systems cryptographically secure pseudorandom number generator.
* `create_keyshares()` and `combine_keyshares()` for distributing key 
  shares to a group using *Shamir's Secret Sharing* algorithm
* `blake2b()` for hashing any R object
* `blake2b_raw()` for hashing raw bytes and strings directly
* `encrypt_obj()` and `decrypt_obj()` for saving encrypted objects to file

## Included Source Code 


The package relies on the cryptographic algorithms supplied by [`monocypher`](https://monocypher.org/)


* x25519 key exchange (Public Key Cryptography)
* RFC 8439 ['Authenticated Encryption with Additional Data (AEAD)'](https://en.wikipedia.org/wiki/Authenticated_encryption#Authenticated_encryption_with_associated_data_(AEAD)) 
i.e.  ChaCha20-Poly1305 combining ChaCha20 stream cipher with 
Poly1305 message authentication code.

Shamir's Secret Sharing uses Daan Sprenkel's [sss](https://github.com/dsprenkels/sss)
code.

## Installation

To install `rmonocypher` from [GitHub](https://github.com/coolbutuseless/rmonocypher) with:

``` r
# install.packages("devtools")
devtools::install_github("coolbutuseless/rmonocypher")
```

# Saving encrypted objects to file

```{r}
file <- tempfile()
encrypt_obj(head(mtcars), filename = file, key = "mykey")
decrypt_obj(filename = file, key = "mykey")
```



# Using the `cryptfile()` connection

Any function which supports reading/writing with a connection, now supports 
automatic encryption/decryption using the `cryptfile()` connection in this package.

Many R functions support connections e.g. `read.csv()`, `saveRDS()`, 
`serialize()`, `png::writePNG()`, and these can all seamlessly use the `cryptfile()` connection.

The output can only be recovered using the same `key` which was 
used for encryption.

```{r}
robj <- head(mtcars)
path <- tempfile()
key  <- argon2("my secret")
saveRDS(robj, cryptfile(path, key))

readRDS(cryptfile(path, key))
```

Attempting to use an incorrect key will not work:

```{r, error=TRUE}
readRDS(cryptfile(path, "wrong key"))
```

Attempting to read the file without decrypting it will not work:

```{r, error = TRUE}
readRDS(path)
```

### More examples using `cryptfile()` connection

```{r eval=FALSE}
# Saving R objects
saveRDS(robj, cryptfile(path, key))
readRDS(cryptfile(path, key))

# CSV files
write.csv(iris, cryptfile(path, key))
read.csv(cryptfile(path, key))

# PNG files
png::writePNG(image, cryptfile(path, key))
png::readPNG(cryptfile(path, key))

# sink()
sink(cryptfile(path, key))
...
sink()

# Text
writeLines(my_diary, cryptfile(path, key))
readLines(cryptfile(path, key))

# cat()
cat(data, cryptfile(path, key))
```



# Simple encrypt/decrypt of raw data and strings

`encrypt()` and `decrypt()` are functions for directly encrypting 
strings and raw vectors

```{r}
library(rmonocypher)

# Plain text. Can be string or raw bytes
dat <- "Hello #RStats"

# Encrypt 
enc <- encrypt(dat, key = "my secret")

# the encrypted data
enc

# Decrypt using the same key
decrypt(enc, key = "my secret", type = 'string')
```

# Encryption keys

The encryption `key` is the core secret information that allows for
encrypting data.

The `key` for encryption may be one of:

* A 32-byte raw vector
* A 64-character hexadecimal string
* A pass-phrase

The `key` may be created by:

* Using random bytes from a cryptographically secure source
* Using Argon2 to derive random bytes from a pass-phrase
* Creating a shared key through key exchange with another person


When calling functions in `{rmonocypher}`, the key may be set explicitly when the function is called, but can also be set
globally for the current session using `options(MONOCYPHER_KEY = "...")`

## Using random bytes as a key

```{r}
# Random raw bytes
key <- as.raw(sample(0:255, 32, TRUE))

# Random raw bytes from rcrypto()
key <- rcrypto(32)

# 64-character hexadecimal string
key <- "82febb63ac2ab2a10193ee40ac711250965ed35dc1ce6a7e213145a6fa753230"
```


## Argon2: Password-based Key Derivation

Argon2 is a resource intensive password-based key derivation scheme.

Use `argon2()` to generate random bytes for keys from a pass-phrase.  It is 
recommended to further defend against attackers using rainbow tables by 
providing extra bytes of `salt`.

If no explicit `salt` is provided, a salt will be derived internally from 
the pass-phrase.  This is deterministic such that the same pass-phase
will always generate the same key.  This is convenient, but not as
secure as using another pass-phrase or random bytes for the `salt`.

```{r echo = FALSE}
set.seed(2024)
```

```{r}
# When no salt is provided, a salt will be 
# derived internally from the pass-phrase.  This is convenient, but 
# not as secure as using random bytes.
argon2("my secret")

# Use text as the salt
argon2("my secret", salt = "salt and vinegar")

# Use a 32-character hexadecimal string as the salt
argon2("my secret", salt = "cefca6aafae5bdbc15977fd56ea7f1eb")

# Use 16-bytes of random data for the salt
argon2("my secret", salt = as.raw(sample(0:255, 16, TRUE)))

# Use 'rcrypto()' to source 16 random bytes for the salt
argon2("my secret", salt = rcrypto(16))
```


## Securely exchange keys over insecure channels with public key encryption.

**Note:** This is an advanced topic, and not essential for regular use of 
encryption when only you are accessing data, or you have a secure way to share
the key with others.

`{rmonocypher}` implement public-key cryptography using x25519. 
X25519 is an elliptic curve Diffie-Hellman key exchange using Curve25519. 
It allows two parties to jointly agree on a shared secret using an insecure channel.

Steps:

1. Both users create a secret key (these are never shared!)
2. Both users derive the public key from their secret key
3. Users swap their public keys. These do not need to be kept secure.
4. Both users use their secret key in conjunction with the other user's
   public key to derive **the exact same key** !
5. Now both users know the same shared key and can encrypt and decrypt 
   messages from each other.

```{r}
# You: Create a secret key and a public key.
# You: Share the public key with other party
your_secret <- argon2("hello")
your_public <- create_public_key(your_secret)

# They: Create a secret key and a public key
# They: Share their public key with you
their_secret <- argon2("goodbye")
their_public <- create_public_key(their_secret)

# You: Use their public key and your secret key
#      to derive the common shared key
create_shared_key(their_public, your_secret)

# They: Use your public key and their secret key
#       to derive the same shared key!
create_shared_key(your_public, their_secret)
```


## Shamir's Secret Sharing

*Shamir's Secret Sharing* algorithm allows a key to be split into multiple parts (*keyshares*)
and shared.  When splitting the key, the number (`k`) is specified to indicate how
many keyshares are required to reconstruct the key.  Any individual *keyshare* cannot reveal the key.  See [wikipedia](https://en.wikipedia.org/wiki/Shamir%27s_secret_sharing).

#### Application example (from wikipedia)

    A company needs to secure their vault. If a single person knows the 
    code to the vault, the code might be lost or unavailable when the 
    vault needs to be opened. If there are several people who know the 
    code, they may not trust each other to always act honestly.

    SSS can be used in this situation to generate shares of the vault's code
    which are distributed to authorized individuals in the company. The 
    minimum threshold and number of shares given to each individual can be 
    selected such that the vault is accessible only by (groups of) authorized 
    individuals. If fewer shares than the threshold are presented, the vault
    cannot be opened.

    By accident, coercion or as an act of opposition, some individuals might 
    present incorrect information for their shares. If the total of correct 
    shares fails to meet the minimum threshold, the vault remains locked.
    
    
```{r}
orig_key <- "337ca9406391140208844c76b536c111f44531adef8d5cebcc68f83ab43cc745"
shares <- create_keyshares(orig_key, n = 6, k = 3)
shares
# Reassemble original key from any 3 keyshares
combine_keyshares(shares[c(1, 2, 3)])
combine_keyshares(shares[c(6, 1, 4)])
```


## Cryptographic Hash `blake2b()` and `blake2b_raw()`

    BLAKE2 is a cryptographic hash function based on BLAKE, created by 
    Jean-Philippe Aumasson, Samuel Neves, Zooko Wilcox-O'Hearn, and Christian 
    Winnerlein. The design goal was to replace the widely used, but broken, 
    MD5 and SHA-1 algorithms in applications requiring high performance in software

For more on why you might want a cryptographic hash vs a regular hash, see
[wikipedia article on cryptographic hash functions](https://en.wikipedia.org/wiki/Cryptographic_hash_function).

```{r}
# Hash of any R object using R's serialization mechanism
blake2b(mtcars)
# Hash of raw vectors and strings directly
blake2b_raw(as.raw(1:20))
blake2b_raw("hello")
```



# Additional data

**Note:** This is an advanced topic, and not essential for regular use of the encryption 
tools in this package.

This package uses ['Authenticated Encryption with Additional Data (AEAD)'](https://en.wikipedia.org/wiki/Authenticated_encryption#Authenticated_encryption_with_associated_data_(AEAD)) 
to encrypt messages.

Additional data (or 'associated data') may be included in the encryption
process. This additional data is part of the message authentication, but 
not part of the encrypted message.

A common way additional data is used is for encrypted data which has an 
unencrypted header containing meta-information.  This header must be readable
before the message is decrypted, but also must not allow tampering.

An example of the use of *additional data* is addressing an encrypted message to a particular
recipient.

The address on the envelope must be readable to allow delivery, but the message
inside needs to remain confidential.  

In this case, the *address* is the *additional data* -
it is sent with the data, but not encrypted.  Because the address forms
part of the message authentication, any modification of the address will 
prevent the authentication of the encrypted payload.

**In the following example**, a message for Judy is encrypted, and the address
on the envelope is used as the *additional data*.

```{r}
# Using additional data to encrypt a message
key      <- argon2("my secret key2")
message  <- 'Meet me in St Louis'
address  <- 'To: Judy'
enc      <- encrypt(message, key, additional_data = address)

# Package the additional data and deliver to recipient
letter <- list(address = address, message = enc)
letter

# Recipient decodes message, and the 'address' forms part of the decryption.
decrypt(letter$message, key = key, type = 'string', additional_data = letter$address)
```

If a malicious third-party tampers with the address, then the message cannot
be authenticated.  E.g. if the letter is altered as if it were being sent
to "Sandra", then decryption will fail:

```{r error = TRUE}
letter$address <- "To: Sandra"
decrypt(letter$message, key = key, type = 'string', additional_data = letter$address)
```




# Technical Notes

* The nonce used within 'monocypher' is 24-bytes (192 bits).  This is large enough that 
  counter/ratcheting mechanisms do not need to be used, and random bytes are 
  unlikely to generate the same nonce twice in any reasonable timeframe.
* The nonce is created internally using random bytes from the cryptographic random number
  generator from the system this is running on.
* There are no magic bytes pre-pended to the data to identify it as encrypted, 
  but the `payload size` field in the file structure is easily identifiable 
  (see below).
* In general when encrypting data using Authenticated Encryption:
    * these should be kept secret:
        * the original data (obviously!)
        * the encryption key.
    * these elements do not need to be kept secret:
        * MAC - message authentication code
        * Number of bytes of data
        * Nonce
        * Salt (if `argon2()` is being used to derive `key` from a pass-phrase)


### File structure

The file structure for encrypted data created by this package is documented here.

The file must hold multiple chunks of streaming encrypted data, each chunk
has a header giving the number of encrypted bytes (payload size) and a MAC (Message Authentication Code).   The `nonce` is stored once at the beginning of the file.


* `[nonce] [frame] [frame] ... [frame]`
    * `[nonce]` = 24 bytes
    * `[frame]` = `[payload size] [mac] [payload]`
        * `[payload size]` = 8 bytes
        * `[mac]` = 16 bytes
        * `[payload]` is a sequence of `payload size` bytes

### Built-in libraries

This package uses the [monocypher](https://monocypher.org) encryption library
v4.0.2 to provide 'authenticated encryption with additional data' (AEAD) 
and Argon2 password-based key derivation.

Shamir's Secret Sharing uses Daan Sprenkel's [sss](https://github.com/dsprenkels/sss)
code.
