---
title: "Decryption outside of R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decryption outside of R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rmonocypher)
```


### Decoding in other programs/languages

The encryption technique used through this package is the XChaCha20-Poly1305 - 
this uses an extended 24-byte nonce.

This may be decoded by any program/language which implements this technique i.e.

* The `standalone-decrypt.c` code included with this package
* The Rust Language `https://docs.rs/chacha20poly1305/latest/chacha20poly1305/`


C sourcecode for decoding outside of R is included in the installed directory
of this package in the file `standalone-decrypt.c`.  

To find it, see the source tarball on CRAN or github, or
when the package is installed in R, find its location using

```{r eval=FALSE}
system.file("standalone-decrypt.c", package = 'rmonocypher', mustWork = TRUE)
```


This file must be compiled with the `monocypher.c/h` source files from [`monocypher`](https://monocypher.org/)

```
gcc -Wall standalone-decrypt.c monocypher.c -o decrypt
./decrypt [filename] [hexadecimal_key] [outfile]
```

